Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-10-29 fstats_pca_lewontin.tex

\documentclass[12pt,fullpage, a4paper]{article}
\usepackage[
margin=1.5cm,
includefoot,
footskip=30pt,
]{geometry}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{comment}
\usepackage[authoryear, comma]{natbib}
\usepackage{lineno}
\usepackage{url}

@@

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As most species, the genetic diversity of human populations is influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, nielsen2020}. Population genetic models use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

In particular, in isolated populations genetic drift will slowly change allele frequencies. As a result, isolated populations are expected to slowly differentiate \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}. However, isolation-by-distance patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.

The interplay of these demographic processes leads to the complex genetic structure observed in present-day human populations \citep{the2015, mallick2016} with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a} that we model in order to reconstruct human demographic history. This is challenging because particularly for large and heterogeneous data sets we cannot expect to devise a single model that captures all processes. A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions. each emphasizing particular aspects of the data.

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}). However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters. For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium. While in most cases these assumptions are violated to some extent and cannot be verified, but we hope that the resulting model fits provide a sufficiently accurate answer for specific research questions.

\paragraph{F-stats}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the data may be unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. I save the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edge lengths measure how much genetic drift has occurred. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, these $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

\paragraph{F-stats and PCA}
The practical issue addressed in this study is how $F$-statistics can be reconciled with one of the most widely used data-driven  techniques, PCA. One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal components, PCs) retaining a maximum of the variance present in the data. In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets. 
%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. 
The PCA-decomposition has been studied for a number of population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}. Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e. as functions of a high-dimensional Euclidean space. For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.

In the next section, I will formally derive the connection between $F$-statistics and PCA, and show how $F$-statistics can be interpreted geometrically, with a particular emphasis on two-dimensional PCA plots. In the results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets.

\section{Theory}
In this section, I will introduce the mathematics and notations for $F$-statistics and PCA. A comprehensive  treatise on PCA is given by e.g. \cite{jolliffe2013} a useful primer on the mathematics is \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

@@

The three $F$-statistics can then be defined as
\begin{subequations}
	\begin{align}
	F_2(X_1, X_2) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	\\
	F_3(X_1; X_2, X_3) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) \\	
	F_4(X_1, X_2; X_3, X_4) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) 	\text{,}
	\end{align}
\end{subequations}
The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
	2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
	2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
	\end{align}
\end{subequations}

As highlighted in the introduction, $F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch between four nodes, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

Thinking of $F$-statistics as branch lengths is useful to understand a number of applications: In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}. One way to do that is using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ denotes an outgroup, and the $X_i$ are a panel of populations that are candidates for the closest match. The  highest values of $F_3$ indicate the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The intuition is given in Figure \ref{fig:outgroupf3}A; where the outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ is highlighted. It represents the length of the branch from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the statistic. In contrast to a simple genetic distance, the sample time has no effect: The branch between $X_U$ and $X_2$, would be shorter than to one between $X_U$ and $X_1$, but the path to the shared junction and hence the $F_3$-statistic would be the same. Larger sets of $F_3$ and $F_4$-statistics are also frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is a violation of treeness. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{patterson2012}. The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A, where (the typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021} showed that these assumptions are not necessary by re-deriving $F$-statistics in a geometric framework. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and that all properties and tests related to treeness can be derived from this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$.

@@

\begin{figure}[!ht]
	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}}
	\label{fig:outgroupf3}
\end{figure}

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I describe here is set up in a way that the similarities to the $F$-statistics framework are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

@@

There are several algorithms that are used to perform PCAs, the most common one is based on singular value decomposition \citep{jolliffe2013}. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
	\begin{equation*}
	y_{il} = x_{il} - \mu_l
	\end{equation*}
	where $\mu_l$ is the mean allele frequency at the $l$-th locus.

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

>>>>

\documentclass[12pt,fullpage, a4paper]{article}
\usepackage[
margin=1.5cm,
includefoot,
footskip=30pt,
]{geometry}
\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{comment}
\usepackage[authoryear, comma]{natbib}
\usepackage{lineno}
\usepackage{url}

@@

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2020}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985a}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{skov2019}.

Particularly for large and heterogeneous data sets, disentangling all these processes is challenging, and we cannot expect to devise a single model catching both broad strokes and minute details of human history. A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions. each emphasizing particular aspects of the data.

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}). However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters. For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium. While in most cases these assumptions are violated to some extent and cannot be verified, we hope that the resulting model fits provide sufficiently accurate answers to specific research questions.

\paragraph{F-statistics}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

\paragraph{F-statistics and PCA}
The practical issue addressed in this study is how $F$-statistics can be reconciled with PCA, one of the most widely used data-driven modeling  techniques. One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal component, PC) retaining a maximum of the variance present in the data. To understand population structure, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize genetic diversity \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets. 
%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. 
The PCA-decomposition has been studied for a number of explicit population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}. Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e. as functions of a high-dimensional Euclidean space. For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.

In the next section, I will formally derive the connection between $F$-statistics and PCA, and show how $F$-statistics can be interpreted geometrically, with a particular emphasis on two-dimensional PCA plots. In the Results section, I will then discuss how some of the most common applications of $F$-statistics manifest themselves on a PCA, and illustrate them on two example data sets, before ending with a discussion.

\section{Theory}
In this section, I will introduce the mathematics and notations for $F$-statistics and PCA. A comprehensive  treatise on PCA is given by e.g. \cite{jolliffe2013}, a useful primer on the mathematics is \cite{pachter2014}, and a helpful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

@@

The three $F$-statistics are defined as
\begin{subequations}
	\begin{align}
	F_2(X_1, X_2) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	\\
	F_3(X_1; X_2, X_3) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) \\	
	F_4(X_1, X_2; X_3, X_4) =& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) 	\text{,}
	\end{align}
\end{subequations}
The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Both $F_3$ and $F_4$ can be written as sums of $F_2$-statistics:
\begin{subequations}
	\begin{align}
	2F_3(X_1; X_2, X_3) &=  F_2(X_1, X_2) + F_2(X_1, X_3) - F_2(X_2, X_3)\label{eq:f3fromf2}\\
	2F_4(X_1, X_2; X_3, X_4) &= F_2(X_1, X_3) + F_2(X_2, X_4) - F_2(X_1,X_4) - F_2(X_2, X_3)\label{eq:f4fromf2}
	\end{align}
\end{subequations}

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

Thinking of $F$-statistics as branch lengths is useful for a number of applications, including building multi-population models \citep{patterson2012, lipson2013}, estimating admixture proportions \citep{petr2019, harney2021} and finding the population most closely related to an unknown sample (``Outgroup''-$F_3$-statistic).

%In particular, one common task is to find the population most closely related to an unknown sample $X_U$ \citep{raghavan2014}. One way to do that is using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ denotes an outgroup, and the $X_i$ are a panel of populations that are candidates for the closest match. The largest value of $F_3$ indicates the population $X_i$ most closely related to $U$, using the outgroup $O$ to correct for differences in sample times. The intuition is given in Figure \ref{fig:outgroupf3}A; where the outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ is highlighted. It represents the length of the branch from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the statistic. In contrast to a simple genetic distance, the sample time has no effect: The branch between $X_U$ and $X_2$, would be shorter than to one between $X_U$ and $X_1$, but the path to the shared junction and hence the $F_3$-statistic would be the same. Larger sets of $F_3$ and $F_4$-statistics are also frequently used for complex models, such as  reconstructing admixture graphs \citep{patterson2012, lipson2013} and estimating admixture proportions \citep{petr2019, harney2021}.

Most commonly however, $F_3$ and $F_4$ are used as tests of treeness \citep{patterson2012}: Negative $F_3$-values correspond to a branch with negative genetic drift, which is not allowed under the null assumption of a tree-like population relationship. Similarly if four populations are related as a tree, then at least one of the $F_4$ statistics between the populations will be zero \citep{buneman1974, patterson2012}.

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021} showed re-deriving $F$-statistics in a geometric framework, showing that these assumptions are not necessary. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. The external branches are independent populations, and so they are expected to drift orthogonally to each other.

@@

\begin{figure}[!ht]
	\includegraphics[width=.5\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}. The circle with a dot denotes a right angle.}
	\label{fig:outgroupf3}
\end{figure}

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis. The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

@@

There are several algorithms that are used to perform PCAs, the most common one is based on singular value decomposition \citep[e.g.][]{jolliffe2013}. In this approach, we first mean-center $\MX$, obtaining a centered matrix $\MY$
	\begin{equation*}
	y_{il} = x_{il} - \mu_l
	\end{equation*}
	where $\mu_l$ is the mean allele frequency at the $l$-th locus.

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations in our sample, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.
